Polynomial phase estimation by phase unwrapping

### 2012-10-19 paper.tex
minor mods

\begin{theorem} \label{thm:asymp_proof} 
Let $\widehat{\mubf}$ be defined by~\eqref{eq:hatmubfLSUdefn} and put $\widehat{\lambdabf}_N = \dealias(\tilde{\mubf} - \widehat{\mubf})$.  Denote the elements of $\widehat{\lambdabf}_N$ by $\widehat{\lambda}_{0,N}, \dots, \widehat{\lambda}_{m,N}$.  Suppose $\Phi_1, \dots, \Phi_N$ are independent and identically distributed with zero intrinsic mean, intrinsic variance $\sigma^2$, and probability density function $f$, then: 
\begin{enumerate}
\item (Strong consistency) $N^k \widehat{\lambda}_{k,N}$ converges almost surely to $0$ as $N\rightarrow\infty$ for all $k = 0, 1, \dots, m$.
\item (Asymptotic normality) If $f(\fracpart{x})$ is continuous at $x = -\nicefrac{1}{2}$ and if $f(-\nicefrac{1}{2}) < 1$ then the distribution of the vector
\[
\left[
\begin{array}
[c]{cccc}%
\sqrt{N} \widehat{\lambda}_{0,N} & N \sqrt{N}\widehat{\lambda}_{1,N}  & \dots & N^m\sqrt{N} \widehat{\lambda}_{m,N}
\end{array}
\right]^\prime
\]
converges to the normal with zero mean and covariance
\[
\frac{\sigma^2}{\left(1-f( -\nicefrac{1}{2}) \right)^{2}} \Cbf^{-1},
\]
where $\Cbf$ is the $m+1$ by $m+1$ Hilbert matrix with elements $C_{ik} = 1/(i + k + 1)$ for $i,k \in \{0, 1, \dots, m\}$.
\end{enumerate}
\end{theorem}
The proof of Theorem~\ref{thm:asymp_proof} is contained within the next two sections. Section~\ref{sec:strongconstproof} proves strong consistency and Section~\ref{sec:centlimitproof} proves asymptotic normality.  Proofs for the case when $m=0$ were given in~\cite{McKilliam_mean_dir_est_sq_arc_length2010} and for the case when $m=1$ were given in~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009}.  The proofs here take a similar approach, but require new techniques.  The proof of asymptotic normality in~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009} contains a subtle error that is amened by the proof given in this paper.
%Before begining the proof we first discuss the assumptions made in the theorem statement.

The theorem gives conditions on $\dealias(\tilde{\mubf} - \widehat{\mubf})$ rather than directly on the difference $\tilde{\mubf} - \widehat{\mubf}$.   To see why this makes sense, consider the case when $m=0$, $\tilde{\mu}_0 = -0.5$ and $\widehat{\mu}_0 = 0.49$, so that $\tilde{\mu}_0 - \widehat{\mu}_0 = -0.99$.  However, the two phases are obviously close, since the phases $\pm 0.5$ are actually the same.  In this case $\dealias(\tilde{\mu}_0 - \widehat{\mu}_0) = \fracpart{\tilde{\mu}_0 - \widehat{\mu}_0} = 0.01$ as expected.  The same reasoning holds for $m > 0$.

@@

The next result we require comes from the calculus of finite differences. For any function $d(n)$ mapping $\reals$ to $\reals$, let 
\[
\Delta_h d(n) = d(n+h) - d(n)
\] 
denote the first difference with interval $h$. Let 
\begin{align*}
\Delta_h^2 d(n) = \Delta_h d(n+h) - \Delta_h d(n) = d(n+2h) - 2d(n+h) + d(n)
\end{align*}
denote the second difference with interval $h$ and similarly let 
\begin{equation}\label{eq:mthdiffformula}
\Delta_h^r d(n) = \Delta_h^{r-1} d(n+h) - \Delta_h^{r-1} d(n) = \sum_{k=0}^{r}\binom{r}{k}(-1)^{r-k}d(n+kh)
\end{equation}
denote the $r$th difference with interval $h$. Since $\sum_{k=0}^{r}\binom{r}{k} = 2^r$ it follows that $\Delta_h^r d(n)$ can be represented by adding and subtracting the 
\[
d(n), \,\, d(n+h), \,\, \dots, \,\, d(n+kh)
\] 
precisely $2^r$ times.

@@

\begin{lemma}\label{eq:convdn}
The distribution of the vector $\dbf_N$, defined in~\eqref{eq:dnandkn}, converges to the multivariate normal with zero mean and covariance matrix $\sigma^2\Cbf$.
\end{lemma}
\begin{proof}
For any constant vector $\alphabf$, let
\[
z_N = \alphabf^\prime \dbf_N = \frac{1}{\sqrt{N}} \sum_{n=1}^{N} \Phi_n \sum_{\ell=0}^{m}\alpha_\ell\left( \frac{n}{N} \right)^\ell.
\]
By Lypanov's central limit theorem $z_N$ is asymptotically normally distributed with zero mean and variance
\[
\lim_{N\rightarrow\infty} \sigma^2 \frac{1}{N} \sum_{n=1}^{N}\left( \sum_{\ell=0}^{m}\alpha_\ell\left( \frac{n}{N} \right)^\ell \right)^2 = \sigma^2 \alphabf^\prime \Cbf \alphabf.
\]
By the Cramer-Wold theorem it follows that $\dbf_N$ is asymptotically normally distributed with zero mean and covariance $\sigma^2 \Cbf$. 
% Recall that the $\ell$th element of $\dbf_N$ is
% \[
% D_{\ell,N} = \frac{1}{\sqrt{N}} \sum_{n=1}^{N}(\tfrac{n}{N})^\ell \Phi_n.
% \]
% Since the $\Phi_1, \dots, \Phi_N$ are independent, each with variance $\sigma^2$, the distribution of $D_{\ell,N}$ converges to the normal with zero mean and variance
% \[
% \lim_{N\rightarrow\infty}\var D_{\ell,N} = \sigma^2 \lim_{N\rightarrow\infty} C_{2\ell,N} = \sigma^2 C_{\ell\ell}
% \]
% by Lypanov's central limit theorem and by using~\eqref{eq:ClkNHilbert}.  The covariance between $D_{\ell,N}$ and $D_{k,N}$ is
% \[
% \covar(D_{\ell,N}, D_{k,N}) = \sigma^2 C_{\ell+k,N} \rightarrow \sigma^2 C_{\ell k}
% \]
% as $N \rightarrow \infty$, again using~\eqref{eq:ClkNHilbert}.  The proof follows.
\end{proof}

@@

Figures~\ref{plot:polyest1}~and~\ref{plot:polyest2} show the sample MSEs obtained for a polynomial phase signal of order $m=3$.  Figure~\ref{plot:polyest1} displays results for the zeroth and first order coefficients $\widehat{\mu}_0$ and $\widehat{\mu}_1$ and Figure~\ref{plot:polyest2} displays results for the second and third order coefficients $\widehat{\mu}_0$ and $\widehat{\mu}_1$.  When $N = 10$ and $50$ the LSU estimator can be computed exactly using a general purpose algorithm for finding nearest lattice points called the sphere~decoder~\cite{Pohst_sphere_decoder_1981,Agrell2002,Viterbo_sphere_decoder_1999}.  This is displayed be the circles in the figures.  When $N=200$ the sphere decoder is computationally intractable and we instead use an approximate nearest point algorithm called the $K$-best method~\cite{Zhan2006_K_best_sphere_decoder}.  This is displayed by the dots.  For the purpose of comparison we have also plotted the results for the $K$-best method when $N = 10$ and $50$.  The asymptotic variance predicted in Theorem~\ref{thm:asymp_proof} is displayed by the dashed line.  Provided the noise variance is small enough (so that the `threshold' is avoided) the sample MSE of the LSU estimator is close to that predicted by Theorem~\ref{thm:asymp_proof}.  The Cram\`{e}r-Rao lower bound for the variance of unbiased polynomial phase estimators in Gaussian noise is also plotted using the solid line~\cite{McKilliam_polycrbs_2012,Peleg1991_CRB_PPS_1991}.  When the noise variance is small the asymptotic variance of the LSU estimator is close to the Cram\`{e}r-Rao lower bound.

@@

This paper has considered the estimation of the coefficients of a noisy polynomial phase signal by least squares phase unwrapping (LSU). It has been shown that the LSU estimator is strongly consistent and asymptotically normally distributed. %Simulations are used to show that the LSU has near maximum likelihood performance in the case that the noise is additive white and Gaussian.  
It is shown in~\cite[Sec~8.1]{McKilliam2010thesis}\cite{McKilliam2009asilomar_polyest_lattice} how the LSU estimator can be computed by finding a nearest lattice point in a particular lattice.  Polynomial time algorithms that compute the nearest point are described in~\cite[Sec.~4.3]{McKilliam2010thesis}, but these are slow algorithms in practice.  A significant outstanding question is whether practically fast nearest point algorithms exist for these specific lattices.  Considering the excellent statistical performance (both theoretically and practically) of the LSU estimator, even fast \emph{approximate} nearest point algorithms are likely to prove useful for the estimation of polynomial phase signals.

>>>>

\begin{theorem} \label{thm:asymp_proof} 
Let $\widehat{\mubf}$ be defined by~\eqref{eq:hatmubfLSUdefn} and put $\widehat{\lambdabf}_N = \dealias(\tilde{\mubf} - \widehat{\mubf})$.  Denote the elements of $\widehat{\lambdabf}_N$ by $\widehat{\lambda}_{0,N}, \dots, \widehat{\lambda}_{m,N}$.  Suppose $\Phi_1, \dots, \Phi_N$ are independent and identically distributed with zero intrinsic mean, intrinsic variance $\sigma^2$, and probability density function $f$, then: 
\begin{enumerate}
\item (Strong consistency) $N^k \widehat{\lambda}_{k,N}$ converges almost surely to $0$ as $N\rightarrow\infty$ for all $k = 0, 1, \dots, m$.
\item (Asymptotic normality) If $f(\fracpart{x})$ is continuous at $x = -\nicefrac{1}{2}$ and if $f(-\nicefrac{1}{2}) < 1$ then the distribution of the vector
\[
\left[
\begin{array}
[c]{cccc}%
\sqrt{N} \widehat{\lambda}_{0,N} & N \sqrt{N}\widehat{\lambda}_{1,N}  & \dots & N^m\sqrt{N} \widehat{\lambda}_{m,N}
\end{array}
\right]^\prime
\]
converges to the normal with zero mean and covariance
\[
\frac{\sigma^2}{\left(1-f( -\nicefrac{1}{2}) \right)^{2}} \Cbf^{-1},
\]
where $\Cbf$ is the $m+1$ by $m+1$ Hilbert matrix with elements $C_{ik} = 1/(i + k + 1)$ for $i,k \in \{0, 1, \dots, m\}$.
\end{enumerate}
\end{theorem}
The proof of Theorem~\ref{thm:asymp_proof} is contained within the next two sections. Section~\ref{sec:strongconstproof} proves strong consistency and Section~\ref{sec:centlimitproof} proves asymptotic normality.  Proofs for the case when $m=0$ were given in~\cite{McKilliam_mean_dir_est_sq_arc_length2010} and for the case when $m=1$ were given in~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009}.  The proofs here take a similar approach, but require new techniques.  %The proof of asymptotic normality in~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009} contains a subtle error that is amended by the proof given in this paper.
%Before begining the proof we first discuss the assumptions made in the theorem statement.
The theorem gives conditions on $\dealias(\tilde{\mubf} - \widehat{\mubf})$ rather than directly on the difference $\tilde{\mubf} - \widehat{\mubf}$.   To see why this makes sense, consider the case when $m=0$, $\tilde{\mu}_0 = -0.5$ and $\widehat{\mu}_0 = 0.49$, so that $\tilde{\mu}_0 - \widehat{\mu}_0 = -0.99$.  However, the two phases are obviously close, since the phases $\pm 0.5$ are actually the same.  In this case $\dealias(\tilde{\mu}_0 - \widehat{\mu}_0) = \fracpart{\tilde{\mu}_0 - \widehat{\mu}_0} = 0.01$ as expected.  The same reasoning holds for $m > 0$.

@@

The next result we require comes from the calculus of finite differences. For any function $d(n)$ mapping $\reals$ to $\reals$, let 
\[
\Delta_h d(n) = d(n+h) - d(n)
\] 
denote the first difference with interval $h$, and let
% \begin{align*}
% \Delta_h^2 d(n) = \Delta_h d(n+h) - \Delta_h d(n) = d(n+2h) - 2d(n+h) + d(n)
% \end{align*}
% denote the second difference with interval $h$ and similarly let 
\begin{equation}\label{eq:mthdiffformula}
\Delta_h^r d(n) = \Delta_h^{r-1} d(n+h) - \Delta_h^{r-1} d(n) = \sum_{k=0}^{r}\binom{r}{k}(-1)^{r-k}d(n+kh)
\end{equation}
denote the $r$th difference with interval $h$. Since $\sum_{k=0}^{r}\binom{r}{k} = 2^r$ it follows that $\Delta_h^r d(n)$ can be represented by adding and subtracting the 
\[
d(n), \,\, d(n+h), \,\, \dots, \,\, d(n+kh)
\] 
precisely $2^r$ times.

@@

\begin{lemma}\label{eq:convdn}
The distribution of the vector $\dbf_N$, defined in~\eqref{eq:dnandkn}, converges to the multivariate normal with zero mean and covariance matrix $\sigma^2\Cbf$.
\end{lemma}
\begin{proof}
For any constant vector $\alphabf$, let
\[
z_N = \alphabf^\prime \dbf_N = \frac{1}{\sqrt{N}} \sum_{n=1}^{N} \Phi_n \sum_{\ell=0}^{m}\alpha_\ell\left( \frac{n}{N} \right)^\ell.
\]
By Lypanov's central limit theorem $z_N$ is asymptotically normally distributed with zero mean and variance
\[
\lim_{N\rightarrow\infty} \sigma^2 \frac{1}{N} \sum_{n=1}^{N}\left( \sum_{\ell=0}^{m}\alpha_\ell\left( \frac{n}{N} \right)^\ell \right)^2 = \sigma^2 \alphabf^\prime \Cbf \alphabf.
\]
By the Cram\`{e}r-Wold theorem it follows that $\dbf_N$ is asymptotically normally distributed with zero mean and covariance $\sigma^2 \Cbf$. 
% Recall that the $\ell$th element of $\dbf_N$ is
% \[
% D_{\ell,N} = \frac{1}{\sqrt{N}} \sum_{n=1}^{N}(\tfrac{n}{N})^\ell \Phi_n.
% \]
% Since the $\Phi_1, \dots, \Phi_N$ are independent, each with variance $\sigma^2$, the distribution of $D_{\ell,N}$ converges to the normal with zero mean and variance
% \[
% \lim_{N\rightarrow\infty}\var D_{\ell,N} = \sigma^2 \lim_{N\rightarrow\infty} C_{2\ell,N} = \sigma^2 C_{\ell\ell}
% \]
% by Lypanov's central limit theorem and by using~\eqref{eq:ClkNHilbert}.  The covariance between $D_{\ell,N}$ and $D_{k,N}$ is
% \[
% \covar(D_{\ell,N}, D_{k,N}) = \sigma^2 C_{\ell+k,N} \rightarrow \sigma^2 C_{\ell k}
% \]
% as $N \rightarrow \infty$, again using~\eqref{eq:ClkNHilbert}.  The proof follows.
\end{proof}

@@

Figures~\ref{plot:polyest1}~and~\ref{plot:polyest2} show the sample MSEs obtained for a polynomial phase signal of order $m=3$.  Figure~\ref{plot:polyest1} displays results for the zeroth and first order coefficients $\widehat{\mu}_0$ and $\widehat{\mu}_1$ and Figure~\ref{plot:polyest2} displays results for the second and third order coefficients $\widehat{\mu}_2$ and $\widehat{\mu}_3$.  When $N = 10$ and $50$ the LSU estimator can be computed exactly using a general purpose algorithm for finding nearest lattice points called the sphere~decoder~\cite{Pohst_sphere_decoder_1981,Agrell2002,Viterbo_sphere_decoder_1999}.  This is displayed by the circles in the figures.  When $N=200$ the sphere decoder is computationally intractable and we instead use an approximate nearest point algorithm called the $K$-best method~\cite{Zhan2006_K_best_sphere_decoder}.  This is displayed by the dots.  For the purpose of comparison we have also plotted the results for the $K$-best method when $N = 10$ and $50$.  
The asymptotic variance predicted in Theorem~\ref{thm:asymp_proof} is displayed by the dashed line.  Provided the noise variance is small enough (so that the `threshold' is avoided) the sample MSE of the LSU estimator is close to that predicted by Theorem~\ref{thm:asymp_proof}.  The Cram\`{e}r-Rao lower bound for the variance of unbiased polynomial phase estimators in Gaussian noise is also plotted using the solid line~\cite{Peleg1991_CRB_PPS_1991}.  When the noise variance is small the asymptotic variance of the LSU estimator is close to the Cram\`{e}r-Rao lower bound.

@@

This paper has considered the estimation of the coefficients of a noisy polynomial phase signal by least squares phase unwrapping (LSU). It has been shown that the LSU estimator is strongly consistent and asymptotically normally distributed. %Simulations are used to show that the LSU has near maximum likelihood performance in the case that the noise is additive white and Gaussian.  
%It is shown in~\cite[Sec~8.1]{McKilliam2010thesis}\cite{McKilliam2009asilomar_polyest_lattice} how the LSU estimator can be computed by finding a nearest lattice point in a particular lattice.  
Polynomial time algorithms that compute the LSU estimator are described in~\cite{McKilliam2010thesis}, but these are slow algorithms in practice.  A significant outstanding question is whether practically fast algorithms exist.  Considering the excellent statistical performance (both theoretically and practically) of the LSU estimator, even fast \emph{approximate} algorithms are likely to prove useful for the estimation of polynomial phase signals.
